Study guide: Analysis of exponential decay models

### 2015-10-18 doc/src/chapters/analysis/main_analysis.tex

>>>>

\section{Various types of errors in a differential equation model}

So far we have been concerned with one type of error, namely the
discretization error committed by replacing the differential equation
problem by a recursive set of difference equations. There are,
however, other types of errors that must be considered too. We can
classify errors into four groups:

\begin{enumerate}
\item model errors

\item data errors

\item discretization errors

\item round-off errors
\end{enumerate}

\noindent
Below, we shall briefly describe and illustrate these four types
of errors.

\subsection{Model errors}

Any mathematical model like $u^{\prime}=-au$, $u(0)=I$, is just an
approximate description of a real-world phenomenon. Suppose a more
accurate model has $a$ as a function of time rather than a
constant. Here we take $a(t)$ as a simple linear function: $a +
pt$. Obviously, $u$ with $p>0$ will go faster to zero with time than a
constant $a$.

The solution of

\[ u^{\prime} = (a + pt)u,\quad u(0)=I,\]
can be shown (see below) to be

\[ u(t) = I e^{t \left(- a - \frac{p t}{2}\right)}\tp\]
With the above $u$ available in a Python function \Verb!u_true(t, I, a, p)!
and the solution from our model $u^{\prime}=-au$ available in \texttt{u(t, I, a)},
we can make some plots of the two models and the error for some values
of $p$. Figure~\ref{decay:analysis:model_errors:fig:model_u} displays
the two curves for $p=0.01, 0.1, 1$, while Figure~\ref{decay:analysis:model_errors:fig:model_e} shows the difference
between the two models as a function of $t$ for the same $p$ values.

\begin{figure}[!ht]  % decay:analysis:model_errors:fig:model_u
  \centerline{\includegraphics[width=1.0\linewidth]{fig-analysis/model_errors_u.pdf}}
  \caption{
  Comparison of two models for three values of $p$. \label{decay:analysis:model_errors:fig:model_u}
  }
\end{figure}
%\clearpage % flush figures decay:analysis:model_errors:fig:model_u

\begin{figure}[!ht]  % decay:analysis:model_errors:fig:model_e
  \centerline{\includegraphics[width=0.8\linewidth]{fig-analysis/model_errors_e.pdf}}
  \caption{
  Discrepancy of Comparison of two models for three values of $p$. \label{decay:analysis:model_errors:fig:model_e}
  }
\end{figure}
%\clearpage % flush figures decay:analysis:model_errors:fig:model_e

The code that was used to produce the plots looks like

\begin{cod}{cbg_blue1}\begin{Verbatim}[numbers=none,fontsize=\fontsize{9pt}{9pt},baselinestretch=0.95,xleftmargin=2mm]
from numpy import linspace, exp
from matplotlib.pyplot import \ 
     plot, show, xlabel, ylabel, legend, savefig, figure, title

def model_errors():
    p_values = [0.01, 0.1, 1]
    a = 1
    I = 1
    t = linspace(0, 4, 101)
    legends = []
    # Work with figure(1) for the discrepancy and figure(2+i)
    # for plotting the model and the true model for p value no i
    for i, p in enumerate(p_values):
        u = model(t, I, a)
        u_true = true_model(t, I, a, p)
        discrepancy = u_true - u
        figure(1)
        plot(t, discrepancy)
        figure(2+i)
        plot(t, u, 'r-', t, u_true, 'b--')
        legends.append('p=%g' % p)
    figure(1)
    legend(legends, loc='lower right')
    savefig('tmp1.png'); savefig('tmp1.pdf')
    for i, p in enumerate(p_values):
        figure(2+i)
        legend(['model', 'true model'])
        title('p=%g' % p)
        savefig('tmp%d.png' % (2+i)); savefig('tmp%d.pdf' % (2+i))
\end{Verbatim}
\end{cod}
\noindent

To derive the analytical solution of the model $u^{\prime}=-(a+pt)u$, $
u(0)=I$, we can use SymPy and the code below. This is somewhat advanced
SymPy use for a newbie, but serves to illustrate the possibilities to
solve differential equations by symbolic software.

\begin{cod}{cbg_blue1}\begin{Verbatim}[numbers=none,fontsize=\fontsize{9pt}{9pt},baselinestretch=0.95,xleftmargin=2mm]
def derive_true_solution():
    import sympy as sym
    u = sym.symbols('u', cls=sym.Function)  # function u(t)
    t, a, p, I = sym.symbols('t a p I', real=True)

    def ode(u, t, a, p):
        """Define ODE: u' = (a + p*t)*u. Return residual."""
        return sym.diff(u, t) + (a + p*t)*u

    eq = ode(u(t), t, a, p)
    s = sym.dsolve(eq)
    # s is sym.Eq object u(t) == expression, we want u = expression,
    # so grab the right-hand side of the equality (Eq obj.)
    u = s.rhs
    print u
    # u contains C1, replace it with a symbol we can fit to
    # the initial condition
    C1 = sym.symbols('C1', real=True)
    u = u.subs('C1', C1)
    print u
    # Initial condition equation
    eq = u.subs(t, 0) - I
    s = sym.solve(eq, C1)  # solve eq wrt C1
    print s
    # s is a list s[0] = ...
    # Replace C1 in u by the solution
    u = u.subs(C1, s[0])
    print 'u:', u
    print sym.latex(u)  # latex formula for reports

    # Consistency check: u must fulfill ODE and initial condition
    print 'ODE is fulfilled:', sym.simplify(ode(u, t, a, p))
    print 'u(0)-I:', sym.simplify(u.subs(t, 0) - I)

    # Convert u expression to Python numerical function
    # (modules='numpy' allows numpy arrays as arguments,
    # we want this for t)
    u_func = sym.lambdify([t, I, a, p], u, modules='numpy')
    return u_func

true_model = derive_true_solution()
\end{Verbatim}
\end{cod}
\noindent

\subsection{Data errors}

\index{Monte Carlo simulation}

By ``data'' we mean all the input parameters to a model, in our case
$I$ and $a$. The values of these may contain errors, or at least
uncertainty. Suppose $I$ and $a$ are measured from some physical
experiments. Ideally, we have many samples of $I$ and $a$ and
from these we can fit probability distributions. Assume that $I$
turns out to be normally distributed with mean 1 and standard deviation 0.2,
while $a$ is uniformly distributed in the interval $[0.5, 1.5]$.

How will the uncertainty in $I$ and $a$ propagate through the model
$u=Ie^{-at}$? That is, what is the uncertainty in $u$ at a particular
time $t$? This answer can easily be answered using \emph{Monte Carlo
simulation}. It means that we draw a lot of samples from the
distributions for $I$ and $a$. For each combination of $I$ and $a$
sample we compute the corresponding $u$ value for selected values of
$t$.  Afterwards, we can for each selected $t$ values make a histogram
of all the computed $u$ values to see what the distribution of $u$
values look like. Figure~\ref{decay:analysis:data_errors:fig} shows the
histograms corresponding to $t=0,1,3$. We see that the distribution of
$u$ values is much like a symmetric normal distribution at $t=0$,
centered around $u=1$. At later times, the distribution gets more
asymmetric and narrower. It means that the uncertainty decreases with
time. From the computed $u$ values we can easily calculate the mean
and standard deviation. The table below shows the mean and standard
deviation values along with the value if we just use the formula
$u=Ie^{-at}$ with the mean values of $I$ and $a$: $I=1$ and $a=1$. As
we see, there is some discrepancy between this latter (naive)
computation and the mean value produced by Monte Carlo simulation.

\begin{center}
\begin{tabular}{llll}
\hline
\multicolumn{1}{c}{ time } & \multicolumn{1}{c}{ mean } & \multicolumn{1}{c}{ st.dev. } & \multicolumn{1}{c}{ $u(t;I=1,a=1)$ } \\
\hline
0    & 1.00 & 0.200   & 1.00           \\
1    & 0.38 & 0.135   & 0.37           \\
3    & 0.07 & 0.060   & 0.14           \\
\hline
\end{tabular}
\end{center}

\noindent
Actually, $u(t;I,a)$ becomes a stochastic variable for each $t$ when
$I$ and $a$ are stochastic variables, as they are in the above
Monte Carlo simulation. The mean of the stochastic $u(t;I,a)$ is
not equal to $u(t;I=1,a=1)$ unless $u$ is linear in $I$ and $a$
(here $u$ is nonlinear in $a$). The accuracy of the Monte Carlo
results increases with increasing number of samples.

\begin{figure}[!ht]  % decay:analysis:data_errors:fig
  \centerline{\includegraphics[width=1.0\linewidth]{fig-analysis/data_errors.pdf}}
  \caption{
  Histogram of solution uncertainty at three time points, due to data errors. \label{decay:analysis:data_errors:fig}
  }
\end{figure}
%\clearpage % flush figures decay:analysis:data_errors:fig

The computer code required to do the Monte Carlo simulation and
produce the plots in Figure~\ref{decay:analysis:data_errors:fig}
is shown below.

\begin{cod}{cbg_blue1}\begin{Verbatim}[numbers=none,fontsize=\fontsize{9pt}{9pt},baselinestretch=0.95,xleftmargin=2mm]
def data_errors():
    from numpy import random, mean, std
    from matplotlib.pyplot import hist
    N = 10000
    # Draw random numbers for I and a
    I_values = random.normal(1, 0.2, N)
    a_values = random.uniform(0.5, 1.5, N)
    # Compute corresponding u values for some t values
    t = [0, 1, 3]
    u_values = {}  # samples for various t values
    u_mean = {}
    u_std = {}
    for t_ in t:
        # Compute u samples corresponding to I and a samples
        u_values[t_] = [model(t_, I, a)
                        for I, a in zip(I_values, a_values)]
        u_mean[t_] = mean(u_values[t_])
        u_std[t_] = std(u_values[t_])

        figure()
        dummy1, bins, dummy2 = hist(
            u_values[t_], bins=30, range=(0, I_values.max()),
            normed=True, facecolor='green')
        #plot(bins)
        title('t=%g' % t_)
        savefig('tmp_%g.png' % t_); savefig('tmp_%g.pdf' % t_)
    # Table of mean and standard deviation values
    print 'time   mean   st.dev.'
    for t_ in t:
        print '%3g    %.2f    %.3f' % (t_, u_mean[t_], u_std[t_])
\end{Verbatim}
\end{cod}
\noindent

\subsection{Discretization errors}

The errors implied by solving the differential equation problem by
the $\theta$-rule has been thoroughly analyzed in the previous
sections. Below are some plots of the error versus time for the
Forward Euler (FE), Backward Euler (BN), and Crank-Nicolson (CN)
schemes for decreasing values of $\Delta t$. Since the difference
in magnitude between the errors in the CN scheme versus the FE ad
BN schemes grows significantly as $\Delta t$ is reduced (the error
goes like $\Delta t^2$ for CN versus $\Delta t$ for FE/BE), we have
plotted the logarithm of the absolute value of the numerical error.

\begin{figure}[!ht]  % decay:analysis:numerical_errors:fig
  \centerline{\includegraphics[width=0.9\linewidth]{fig-analysis/numerical_errors.pdf}}
  \caption{
  Discretization errors in various schemes for 4 time step values. \label{decay:analysis:numerical_errors:fig}
  }
\end{figure}
%\clearpage % flush figures decay:analysis:numerical_errors:fig

The computer code used to generate the plots appear next. It makes use
of a \texttt{solver} function
for computing the numerical solution of $u^{\prime}=-au$ with
the $\theta$-rule.

\begin{cod}{cbg_blue1}\begin{Verbatim}[numbers=none,fontsize=\fontsize{9pt}{9pt},baselinestretch=0.95,xleftmargin=2mm]
def discretization_errors():
    from numpy import log, abs
    I = 1
    a = 1
    T = 4
    t = linspace(0, T, 101)
    schemes = {'FE': 0, 'BE': 1, 'CN': 0.5}  # theta to scheme name
    dt_values = [0.8, 0.4, 0.1, 0.01]
    for dt in dt_values:
        figure()
        legends = []
        for scheme in schemes:
            theta = schemes[scheme]
            u, t = solver(I, a, T, dt, theta)
            u_e = model(t, I, a)
            error = u_e - u
            # Plot log(error), but exclude error[0] since it is 0
            plot(t[1:], log(abs(error[1:])))
            legends.append(scheme)
        xlabel('t');  ylabel('log(abs(numerical error))')
        legend(legends, loc='upper right')
        title(r'$\Delta t=%g$' % dt)
        savefig('tmp_dt%g.png' % dt); savefig('tmp_dt%g.pdf' % dt)
\end{Verbatim}
\end{cod}
\noindent

\subsection{Rounding errors}

Real numbers on a computer are represented by \href{{https://en.wikipedia.org/wiki/Floating_point}}{floating-point numbers}, which means that just a finite
number of digits are stored and used. Therefore, the floating-point number
is an approximation to the underlying real number. When doing
arithmetics with floating-point numbers, there will be small
approximation errors, called round-off errors or rounding errors,
that may or may not accumulate in comprehensive computations.
A typical example is shown below.

\begin{cod}{cbg_blue1}\begin{Verbatim}[numbers=none,fontsize=\fontsize{9pt}{9pt},baselinestretch=0.95,xleftmargin=2mm]
>>> 1.0/51*51
1.0
>>> 1.0/49*49
0.9999999999999999
\end{Verbatim}
\end{cod}
\noindent
We see there is not an exact result in the latter case, but a rounding
error of $10^{-16}$. This is the typical level of a rounding error
from an arithmetic operation with 64 bit floating-point numbers
(\texttt{float} object in Python, often called \texttt{double} or double precision
in other languages).

What is the effect of using \texttt{float} objects and not exact arithmetics?
Python has a \texttt{Decimal} object in the \href{{https://docs.python.org/2/library/decimal.html}}{\nolinkurl{decimal}} module that allows us
to use as many digits in floating-point numbers as we like. We take
1000 digits as the true answer where rounding errors are negligible,
and then we run our numerical algorithm (the Crank-Nicolson scheme to
be precise) with \texttt{Decimal} objects for all real numbers and test the
error arising from using 4, 16, 64, and 128 digits.

When computing with numbers around unity in size, we typically
get a rounding error of $10^{-d}$, where $d$ is the number
of digits used. However, if we compute with numbers that
are larger, e.g., the $u$ values implied by $I=1000$ and $a=100$,
the rounding errors increase to about $10^{-d+3}$. Below is
a table of the the computed maximum rounding error for various number of digits
and two different magnitudes of $I$ and $a$.

\begin{center}
\begin{tabular}{rll}
\hline
\multicolumn{1}{c}{ digits } & \multicolumn{1}{c}{ $I=1$, $a=1$ } & \multicolumn{1}{c}{ $I=1000$, $a=100$ } \\
\hline
4      & $3.05\cdot 10^{-4}$   & $3.05\cdot 10^{-1}$   \\
16     & $1.71\cdot 10^{-16}$  & $1.58\cdot 10^{-13}$  \\
64     & $2.99\cdot 10^{-64}$  & $2.06\cdot 10^{-61}$  \\
128    & $1.60\cdot 10^{-128}$ & $2.41\cdot 10^{-125}$ \\
\hline
\end{tabular}
\end{center}

\noindent
The computer code for doing these experiments need a new version
of the \texttt{solver} function where we do arithmetics with \texttt{Decimal}
objects:

\begin{cod}{cbg_blue1}\begin{Verbatim}[numbers=none,fontsize=\fontsize{9pt}{9pt},baselinestretch=0.95,xleftmargin=2mm]
def solver_decimal(I, a, T, dt, theta):
    """Solve u'=-a*u, u(0)=I, for t in (0,T] with steps of dt."""
    from numpy import zeros, linspace
    from decimal import Decimal as D
    dt = D(dt)
    a = D(a)
    theta = D(theta)
    Nt = int(round(D(T)/dt))
    T = Nt*dt
    u = zeros(Nt+1, dtype=object)  # array of Decimal objects
    t = linspace(0, float(T), Nt+1)

    u[0] = D(I)               # assign initial condition
    for n in range(0, Nt):    # n=0,1,...,Nt-1
        u[n+1] = (1 - (1-theta)*a*dt)/(1 + theta*dt*a)*u[n]
    return u, t
\end{Verbatim}
\end{cod}
\noindent
The function below carries out the experiments. Note that we can
set the number of digits as we want through the \texttt{decimal.getcontext().prec}
variable.

\begin{cod}{cbg_blue1}\begin{Verbatim}[numbers=none,fontsize=\fontsize{9pt}{9pt},baselinestretch=0.95,xleftmargin=2mm]
def rounding_errors(I=1, a=1, T=4, dt=0.1):
    import decimal
    from numpy import log, array, abs
    digits_values = [4, 16, 64, 128]
    # "Exact" arithmetics is taken as 1000 decimals here
    decimal.getcontext().prec = 1000
    u_e, t = solver_decimal(I=I, a=a, T=T, dt=dt, theta=0.5)
    for digits in digits_values:
        decimal.getcontext().prec = digits  # set no of digits
        u, t = solver_decimal(I=I, a=a, T=T, dt=dt, theta=0.5)
        error = u_e - u
        error = array(error[1:], dtype=float)
        print '%d digits, max abs(error): %.2E' % \ 
              (digits, abs(error).max())
\end{Verbatim}
\end{cod}
\noindent

\subsection{Discussion of the size of various errors}
